---
title: "SparCC (global gut full data)"
output:
  html_document:
    theme: united
    fig_width: 6
    fig_height: 6
  pdf_document:
    fig_width: 6
    fig_height: 6
---
Back to [Table of Contents](index.html)  

**All of the code in this page is meant to be run on the command line unless otherwise specified.**

##Install SparCC
Download the repository from the [downloads page](https://bitbucket.org/yonatanf/sparcc/downloads). Unzip the folder, and put it somewhere convenient (like in your course code repository folder). Test that it is working by running the following command on the command line:


```{r eval=TRUE, engine='bash'}
# Test that SparCC is running. 
# Note that you will have to fix the path to be the correct path 
# to the "SparCC.py" file on your computer.
python ../../../sparcc/SparCC.py -h

```

## Prepare data
We will run SparCC on the Global Gut genus data, only including adults living in the USA. We will also choose a subset of the more prevalent genera (present in about 20% of people or more) for testing to keep things running quickly.
```{r eval=TRUE, echo=FALSE, engine='bash'}
rm -rf taxa-USA-adults
```

```{r eval=TRUE, engine='bash'}
# RUN ON THE COMMAND LINE

# First, extract only adults living in the USA.
filter_samples_from_otu_table.py -i otu_table.biom -m map.txt -o otu_table_USA_adults.biom -s "AGE_GROUP:ADULT;COUNTRY:GAZ:United States of America" --output_mapping_fp map_USA_adults.txt

# Summarize taxa at the genus level
summarize_taxa.py -i otu_table_USA_adults.biom -L 6 -o taxa-USA-adults

# remove genera present in < 60 samples
filter_otus_from_otu_table.py -i taxa-USA-adults/otu_table_USA_adults_L6.biom -s 60 -o taxa-USA-adults/otu_table_USA_adults_L6_s60.biom

# create a text version (for SparCC) and a JSON version (for R)
biom convert -i taxa-USA-adults/otu_table_USA_adults_L6_s60.biom --to-json -o taxa-USA-adults/otu_table_USA_adults_L6_s60_json.biom
biom convert -i taxa-USA-adults/otu_table_USA_adults_L6_s60.biom --to-tsv -o taxa-USA-adults/otu_table_USA_adults_L6_s60.txt

# Now remove the first line of the taxon file. Same would apply to an OTU table.
sed 1d taxa-USA-adults/otu_table_USA_adults_L6_s60.txt > taxa-USA-adults/otu_table_USA_adults_L6_s60_for_sparcc.txt

```

## Run SparCC
```{r eval=TRUE, engine='bash'}
python ../../../sparcc/SparCC.py taxa-USA-adults/otu_table_USA_adults_L6_s60_for_sparcc.txt
```

The correlation output will be in the file cor_mat_SparCC.out.

## Comparison to Spearman correlation

Now we will compare SparCC to Spearman correlation in R.

**The following commands will be run in R.**

First, load the data into **R**.
```{r eval=TRUE}
# load the biom library
library('biom')

# read in the biom table and extract the data
x <- t(as.matrix(biom_data(read_biom('taxa-USA-adults/otu_table_USA_adults_L6_s60_json.biom'))))

# read in the sparcc results
sparcc.mat <- read.table('cor_mat_SparCC.out',sep='\t',head=T,row=1)

```

Plot a histogram of correlations inferred by Spearman correlation.
```{r eval=TRUE}
cor.mat <- cor(x,method='spear')
hist(cor.mat[upper.tri(cor.mat)], br=30, xlim=c(-1,1))
```

There are a number of large correlations found above 0.5. Let us compare to SparCC.
```{r eval=TRUE}
hist(sparcc.mat[upper.tri(sparcc.mat)],br=6,xlim=c(-1,1))
```

SparCC found no large correlations, indicating that the correlations found by Spearman correlation could be explained by compositional artifacts.

## Does SparCC fail when there are many more samples than features?

In the previous example, there were around 250 samples and only 40 genera because we kept only highly prevalence genera. Perhaps the fact that there were so many more samples than features caused SparCC to underestimate the true correlations. This could happen because it is assuming most correlations are close to zero, and the more samples there are the more freedom it has to find ways in which different sets of absolute abundances could have made the correlations close to zero. Therefore let's take just the first 30 samples and rerun:

**To be run on the command line**.
```{r eval=TRUE, engine='bash'}
# RUN ON THE COMMAND LINE

# Cut just the first 30 subjects from the USA adults (columns 1-31)
cut -f 1-31 taxa-USA-adults/otu_table_USA_adults_L6_s60_for_sparcc.txt > taxa-USA-adults/otu_table_USA_adults_L6_s60_30subjects_for_sparcc.txt

# run SparCC again
python ../../../sparcc/SparCC.py taxa-USA-adults/otu_table_USA_adults_L6_s60_30subjects_for_sparcc.txt -c cor_mat_30subjects_SparCC.out
```

Now plot the histogram again in R.

**To be run in R**.

```{r eval=TRUE}
# RUN IN R

# read in the sparcc results using only 30 subjects
sparcc.mat <- read.table('cor_mat_30subjects_SparCC.out',sep='\t',head=T,row=1)
cor.mat <- cor(x,method='spear')
hist(sparcc.mat[upper.tri(sparcc.mat)], br=10, xlim=c(-1,1))
```

Here we see that although the correlations are still much smaller than the Spearman correlations, they are larger than with the 250-sample data. We can also keep the 250 samples but also keep the full genus table (308 genera):

**To be run on the command line**.
```{r eval=TRUE, engine='bash'}
# RUN ON THE COMMAND LINE

# Convert the full genus-level table to SparCC format
sed 1d taxa-USA-adults/otu_table_USA_adults_L6.txt > taxa-USA-adults/otu_table_USA_adults_L6_for_sparcc.txt

# run SparCC again
python ../../../sparcc/SparCC.py taxa-USA-adults/otu_table_USA_adults_L6_for_sparcc.txt -c cor_mat_all_genera_SparCC.out
```

Now plot the histogram again in R.

**To be run in R**.

```{r eval=TRUE}
# RUN IN R

# read in the sparcc results using only 30 subjects
sparcc.mat <- read.table('cor_mat_all_genera_SparCC.out',sep='\t',head=T,row=1)
cor.mat <- cor(x,method='spear')
hist(sparcc.mat[upper.tri(sparcc.mat)], br=10, xlim=c(-1,1))
```

