---
title: "Loading Microbiome Data"
output:
  html_document:
    theme: united
---
Back to [Table of Contents](index.html)  

After you have [installed QIIME](01.installing.qiime.html) and [generated an OTU table](02.first.steps.with.qiime.html), you can load the OTU table into R.

**Install biom package**
```{r eval=FALSE}
install.packages('biom',repo='http://cran.wustl.edu')
```

**Load biom package**
```{r}
library('biom')
```

**Convert BIOM file to JSON format**  
If you have data in a "new" BIOM format (HDF5), you first need to convert to JSON format first.  
```{r eval=FALSE, engine='bash'}
# make a JSON-formatted OTU table for loading into R
biom convert -i otus_closed_ref/otu_table.biom -o otu_table_json.biom --to-json
```

**Load global gut data using biom package**
```{r eval=TRUE}
gg.otus.biom <- read_biom('../data/globalgut/otu_table_json.biom')
```

**Extract data matrix (OTU counts) from biom table**
```{r eval=TRUE}
gg.otus <- as.matrix(biom_data(gg.otus.biom))

# transpose so that rows are samples and columns are OTUs
gg.otus <- t(gg.otus)
```

**Plot histogram of sample depths**
```{r}
depths <- rowSums(gg.otus)
hist(depths,breaks=30)
```

**Plot histogram of OTU frequencies**
```{r}
otu.counts <- colSums(gg.otus > 0)
hist(otu.counts,breaks=30)
```

**Remove OTUs present in < 10% of samples**
```{r}
gg.otus <- gg.otus[,colMeans(gg.otus > 0) >= .1]
depths <- rowSums(gg.otus)
dim(gg.otus)
```

**Re-plot histogram of OTU frequencies now that we removed singletons**
```{r}
otu.counts <- colSums(gg.otus > 0)
hist(otu.counts,breaks=30)
```

**Remove any samples with very low depth**
```{r}
sort(depths)[1:10]
gg.otus <- gg.otus[depths >= 1000,]
dim(gg.otus)
```

**Load mapping file**
```{r}
gg.map <- read.table('../data/globalgut/map.txt',sep='\t',head=T,row=1,check=F,comment='')

```

**Ensure that mapping file and OTU table contain the sample samples in the same order**
```{r}
sample.ids <- intersect(rownames(gg.otus), rownames(gg.map))

# might as well put the samples in alphabetical order
sample.ids <- sort(sample.ids)

# in R you can subset using sample IDs or numerical indices. Most languages only use indices.
gg.otus <- gg.otus[sample.ids,]
gg.map <- gg.map[sample.ids,]
dim(gg.otus)
dim(gg.map)
```