---
title: "Day 8: Beta Diversity (Guerrero Negro)"
output:
  html_document:
    theme: united
    fig_width: 6
    fig_height: 6
  pdf_document:
    fig_width: 6
    fig_height: 6
---
Back to [Table of Contents](../../doc/index.html)  

**All of the code in this page is meant to be run in ```R``` unless otherwise specified.**

Install biom package and vegan package if not installed.
```{r eval=FALSE}
install.packages(c('biom','vegan'),repo='http://cran.wustl.edu')
```

Load biom package, load data
```{r, eval=TRUE}
library('biom')
library('vegan')

# load biom file
otus.biom <- read_biom('otu_table_json.biom')

# Extract data matrix (OTU counts) from biom table
otus <- as.matrix(biom_data(otus.biom))

# transpose so that rows are samples and columns are OTUs
otus <- t(otus)

# load mapping file
map <- read.table('map.txt', sep='\t', comment='', head=T, row.names=1)
```

It is extremely important to ensure that your OTU table and metadata table sample IDs are lined up correctly.
```{r, eval=TRUE}
# see rownames of map and otus
rownames(map)
rownames(otus)

# find the overlap
common.ids <- intersect(rownames(map), rownames(otus))

# get just the overlapping samples
otus <- otus[common.ids,]
map <- map[common.ids,]
```

See dimensions of OTU table
```{r}
dim(otus)
```

See dimensions of mapping file
```{r}
dim(map)
```


Get five different distances metrics
```{r, eval=TRUE}

# get Euclidean distance
d.euc <- dist(otus)

# get Bray-Curtis distances (default for Vegan)
d.bray <- vegdist(otus)

# get Chi-square distances using vegan command
# we will extract chi-square distances from correspondence analysis
my.ca <- cca(otus)
d.chisq <- as.matrix(dist(my.ca$CA$u[,1:2]))

# load unweighted and weighted unifrac
d.uuf <- read.table('beta/unweighted_unifrac_otu_table.txt', sep='\t',head=T,row=1)
d.wuf <- read.table('beta/weighted_unifrac_otu_table.txt', sep='\t',head=T,row=1)

# ensure that these last two matrices have the same samples in the 
# same order as the metadata table
d.uuf <- d.uuf[common.ids, common.ids]
d.wuf <- d.wuf[common.ids, common.ids]
```

Now run principal coordinates embedding on the distance metrics
```{r, eval=TRUE}

# Run PCoA (not PCA)
pc.euc <- cmdscale(d.euc, k=2)

# Bray-Curtis principal coords
pc.bray <- cmdscale(d.bray,k=2)

# get first two dimensions of chi-square coordinates:
pc.chisq <- my.ca$CA$u[,1:2]

# get first two dimensions of unifrac distances:
pc.uuf <- cmdscale(d.uuf, k=2)
pc.wuf <- cmdscale(d.wuf, k=2)
```

Plot Euclidean distances with gradient colors
```{r, eval=TRUE}
# makes a gradient from red to blue
my.colors <- colorRampPalette(c('red','blue'))(10)

# plot Euclidean PCoA coords using color gradient
# based on layer (1...10)
layer <- map[,'LAYER']
plot(pc.euc[,1], pc.euc[,2], col=my.colors[layer], cex=3, pch=16)
```

Plot Bray-Curtis distances with gradient colors
```{r, eval=TRUE}

# Plot Bray-Curtis PCoA
plot(pc.bray[,1], pc.bray[,2], col=my.colors[layer], cex=3, pch=16)
```

Plot Chi-square distances with gradient colors
```{r, eval=TRUE}

# Plot Chi-square PCoA
plot(pc.chisq[,1], pc.chisq[,2], col=my.colors[layer], cex=3, pch=16)
```

Plot unweighted UniFrac distances with gradient colors
```{r, eval=TRUE}

plot(pc.uuf[,1], pc.uuf[,2], col=my.colors[layer], cex=3, pch=16)
```

Plot weighted UniFrac distances with gradient colors
```{r, eval=TRUE}

plot(pc.wuf[,1], pc.wuf[,2], col=my.colors[layer], cex=3, pch=16)
```

Note: to make a PDF:
```{r, eval=FALSE}
pdf("chisq.pdf",width=5,height=5)
plot(pc.chisq[,1], pc.chisq[,2], col=my.colors[map[,'LAYER']], cex=3, pch=16)
dev.off()
```

Let's plot pairwise comparisons of the different distance metrics
```{r eval=TRUE}
d.vector.matrix <- cbind(as.numeric(d.euc), as.numeric(d.bray), as.numeric(as.dist(d.chisq)), as.numeric(as.dist(as.matrix(d.uuf))), as.numeric(as.dist(as.matrix(d.wuf))))
colnames(d.vector.matrix) <- c('Euc','BC','ChiSq','UUF','WUF')
pairs(d.vector.matrix)
```

And plot the pairwise pearson correlations
```{r eval=TRUE}
cor(d.vector.matrix)
```

